\begin{thebibliography}{99}

    \addcontentsline{toc}{part}{Bibliography}
    
    \bibitem{MiguelRapado} M.A. Rapado Tamarit, B. Moreno Santamar\'ia, \& J.A. Hern\'andez Ramos. Programming with Visual Studio: Fortran \& Python \& C++. 
                          Amazon Kindle Direct Publishing 2020. 

    \bibitem{Trefethen} Lloyd N. Trefethen \&  D. Bau. Numerical linear algebra. SIAM 1977.
    
    \bibitem{NumericalRecipes} William H. Press, Saul A. Teukolsky, William T. Vetterling \& Brian P. Flannery. Numerical Recipes
    in Fortran 77,
    The Art of Scientific Computing
    Second Edition. Cambridge University Press 1992.
    
    \bibitem{HairerI} E. Hairer, S. P. Norsett, G.Wanner. Solving Ordinary
    Differential Equations I: Nonstiff Problems, Second Revised Edition. Springer Series in Computational Mathematics 2008.
    
    \bibitem{HairerII} E. Hairer, C. Lubich, G.Wanner. Solving Ordinary
    Differential Equations II: Stiff and Differential-Algebraic Problems, Second Revised Edition. Springer Series in Computational Mathematics 1996.
    
%    \bibitem{HairerIII} E. Hairer, C. Lubich, G.Wanner. Geometric Numerical Integration
%    Structure-Preserving Algorithms for Ordinary Differential Equations, Second Revised Edition. Springer Series in Computational Mathematics 2006.
    
%    \bibitem{PracticalAstrodynamics} A. de Iaco. Practical Astrodynamics. Springer Aerospace Technology 2018.
    
    \bibitem{Shampine} L. Shampine. Numerical solution of ordinary differential equations. Chapman \& Hall 1994.
    
    \bibitem{JAHR} J.A. Hernandez Ramos. C\'alculo num\'erico en Ecuaciones Diferenciales Ordinarias.
                  Second Edition Amazon Kindle Direct Publishing 2020.
    
    \bibitem{JAHRFortran} J.A. Hernandez Ramos \& M. Zamecnik Barros. FORTRAN 95, programación multicapa para la simulación de sistemas físicos. Second Edition Amazon Kindle Direct Publishing 2019.
    
    \bibitem{Dormand} J.R.Dormand \& P.J.Prince. A family of embedded Runge-Kutta formulae. Journal of Computational and Applied Mathematics
    Volume 6, Issue 1, March 1980, Pages 19-26.

    \bibitem{Prince} J.R.Dormand \& P.J.Prince. High order embedded Runge-Kutta formulae. Journal of Computational and Applied Mathematics
    Volume 7, Issue 1, March 1981, Pages 67-75.
    
%    \bibitem{Mercury} C. Körber et al. A primer to numerical simulations: The perihelion motion of Mercury. Institut für Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4),
%    Forschungszentrum Jülich, D-52425 Jülich, Germany.

     \bibitem{HenonHeiles} Euaggelos E. Zotos. Classifying orbits in the classical Henon-Heiles Hamiltonian system. Nonlinear Dynamics (NODY), 2015, vol. 79, pp. 1665-1677. 
     
%    \bibitem{Effective} J. Wells. Effective Theories
%    in Physics: From Planetary Orbits to Elementary Particle Masses. Springer, 2012.
%    
%    \bibitem{CERN} J. Wells. When effective theories predict: the inevitability of Mercury’s anomalous perihelion precession. CERN Theoretical Physics. Department of Physics, University of Michigan. November , 2016
    
    \bibitem{Lambert} J. D. Lambert. Numerical Methods for ordinary Differential systems: The Initial Value Problem. John Wiley \& Sons. May, 1993.
    
    
    
\end{thebibliography}
